---
title: "pandas / polars"
subtitle: "Lecture 10"
author: "Dr. Colin Rundel"
footer: "Sta 663 - Spring 2026"
format:
  revealjs:
    theme: slides.scss
    transition: fade
    slide-number: true
    self-contained: true
execute:
  echo: true
---

```{r setup, message=FALSE, warning=FALSE, include=FALSE}
{{< include _setup.R >}}
```

```{python, include=FALSE}
import numpy as np
import pandas as pd

import matplotlib.pyplot as plt

np.set_printoptions(edgeitems=3, linewidth=180)
```

```{python}
#| include: false

np.random.seed(42)

n = 5
d = {
  "id":     np.random.randint(100, 999, n),
  "weight": np.random.normal(70, 20, n),
  "height": np.random.normal(170, 15, n),
  "date":   pd.date_range(start='2/1/2025', periods=n, freq='D')
}

df = pd.DataFrame(d, index = (["anna","bob","carol", "dave", "erin"]))
```


## Pivoting - long to wide 

::: {.small}
```{python include=FALSE}
df = pd.DataFrame({
   "country": ["A","A","A","A","B","B","B","B","C","C","C","C"],
    "year":   [1999,1999,2000,2000,1999,1999,2000,2000,1999,1999,2000,2000],
    "type":   ["cases","pop","cases","pop","cases","pop","cases","pop","cases","pop","cases","pop"],
    "count":  ["0.7K", "19M", "2K", "20M", "37K", "172M", " 80K", "174M", "212K", "1T", "213K", "1T"]
})
```
:::

:::: {.columns .small}
::: {.column width='50%'}
```{python}
df
```
:::

::: {.column width='50%' .fragment}
```{python}
df_wide = df.pivot(
  index=["country","year"], 
  columns="type", 
  values="count"
)
df_wide
```
:::
::::

## pivot indexes

:::: {.columns .small}
::: {.column width='50%'}
```{python}
df_wide.index
df_wide.columns
```
:::

::: {.column width='50%' .fragment}
```{python}
( df_wide
  .reset_index()
  .rename_axis(
    columns=None
  )
)
```
:::
::::


## Wide to long (melt)

```{python include=FALSE}
df = pd.DataFrame({
  "country": ["A","B","C"],
  "1999":    ["0.7K","37K","212K"],
  "2000":    ["2K","80K","213K"]
})
```

:::: {.columns .small}
::: {.column width='50%'}
```{python}
df
```
:::

::: {.column width='50%' .fragment}
```{python}
df_long = df.melt(
  id_vars="country", 
  var_name="year",
  value_name="value"
)
df_long
```
:::
::::


## Exercise 1 - Tidying

How would you tidy the following data frame so that the rate column is split into cases and population columns?

::: {.small}
```{python}
df = pd.DataFrame({
  "country": ["A","A","B","B","C","C"],
  "year":    [1999, 2000, 1999, 2000, 1999, 2000],
  "rate":    ["0.7K/19M", "2K/20M", "37K/172M", "80K/174M", "212K/1T", "213K/1T"]
})
df
```
:::




# Split-Apply-Combine

```{r include=FALSE}
library(tidyverse)
d = readr::read_csv(
  "https://raw.githubusercontent.com/UBC-MDS/programming-in-python-for-data-science/master/data/cereal.csv"
)

d %>%
  mutate(
    mfr = case_when(
      mfr == "A" ~ "Maltex",
      mfr == "G" ~ "General Mills",
      mfr == "K" ~ "Kellogg's",
      mfr == "N" ~ "Nabisco",
      mfr == "P" ~ "Post",
      mfr == "Q" ~ "Quaker Oats",
      mfr == "R" ~ "Ralston Purina"
    )
  ) %>%
  select(-sodium, -potass, -vitamins, -shelf, -weight, -cups) %>%
  select(-(protein:carbo)) %>%
  write_csv("data/cereal.csv")
```

## cereal data 

::: {.xsmall}
```{python}
cereal = pd.read_csv("https://sta663-sp26.github.io/slides/data/cereal.csv"); cereal
```
:::

::: {.aside}
From [UBC-MDS/programming-in-python-for-data-science](https://raw.githubusercontent.com/UBC-MDS/programming-in-python-for-data-science/master/data/)
:::

## groupby

Groups can be created within a DataFrame via `groupby()` - these groups are then used by the standard aggregation methods (e.g. `sum()`, `mean()`, `std()`, etc.).

::: {.small}
```{python}
cereal.groupby("type")
cereal.groupby("type").groups
cereal.groupby("mfr").groups
```
:::

## groupby and aggregation methods

::: {.small}
```{python}
#| error: true
cereal.groupby("type").mean()
```
:::

. . .

:::: {.columns .small}
::: {.column width='50%'}
```{python error=TRUE}
( cereal
  .groupby("type")
  .mean(numeric_only=True)
)
```
:::

::: {.column width='50%' .fragment}
```{python error=TRUE}
cereal.groupby("mfr").size()
```
:::
::::

## Selecting groups

Groups can be accessed via `get_group()` 

::: {.xsmall}
```{python error=TRUE}
cereal.groupby("type").get_group("Hot")

cereal.groupby("mfr").get_group("Post")
```
:::

## Iterating over groups

`DataFrameGroupBy` objects can also be iterated over:

::: {.xxsmall}
```{python}
for name, group in cereal.groupby("type"):
  print(f"# {name}\n{group}\n\n")
```
:::



## Aggregation


The `aggregate()` (or `agg()`) method can be used to compute summary statistics for each group.

::: {.small}
```{python}
#| error: true
cereal.groupby("mfr").agg("mean")
```
:::

. . .

::: {.small}
```{python}
#| error: true
cereal.groupby("mfr").agg("mean", numeric_only = True)
```
:::

::: {.aside}
Think `summarize()` from dplyr.
:::


## Aggregation by column

::: {.small}
```{python}
cereal.groupby("mfr").agg({
  "calories": ['min', 'max'],
  "sugars":   ['median'],
  "rating":   ['sum', 'count']
})
```
:::


::: {.aside}
Note that this results in a DataFrame with a column multiindex.
:::

## Named aggregation

It is also possible to use special syntax to aggregate specific columns into a named output column,

::: {.xsmall}
```{python}
cereal.groupby("mfr", as_index=False).agg(
  min_cal = ("calories", "min"),
  max_cal = ("calories", max),
  med_sugar = ("sugars", "median"),
  avg_rating = ("rating", np.mean)
)
```
:::

::: {.aside}
Tuples can also be passed using `pd.NamedAgg()` but this offers no additional functionality.
:::


## Transformation

The `transform()` method returns a DataFrame with the aggregated result matching the size (or length 1) of the input group(s),

:::: {.columns .xsmall}
::: {.column width='50%'}
```{python}
#| error: true
( cereal
  .groupby("mfr")
  .transform(
    np.mean, numeric_only=True
  )
)
```
:::

::: {.column width='50%' .fragment}
```{python}
( cereal
  .groupby("type")
  .transform(
    "mean", numeric_only=True
  )
)
```
:::
::::


## Practical transformation

`transform()` will generally be most useful via a user-defined function; the lambda is applied to each column of each group.

::: {.xsmall}
```{python}
( cereal
  .drop(["name","type"], axis=1)
  .groupby("mfr")
  .transform( lambda x: (x - np.mean(x))/np.std(x, axis=0) ) 
)
```
:::


## Filtering groups {.scrollable}

::: {.small}
`filter()` also respects groups and allows for the inclusion / exclusion of groups based on user-specified criteria.
:::

::: {.panel-tabset}

### filter 
::: {.xxsmall}
```{python}
( cereal
  .groupby("mfr")
  .filter(lambda x: len(x) > 10)
)
```
:::

### Group sizes

::: {.columns .xsmall}
```{python}
( cereal
  .groupby("mfr")
  .size()
)
```
:::


:::

# ![](imgs/polars_github_banner.svg){fig-align="center"}

## polars

> Polars is a blazingly fast DataFrame library for manipulating structured data. The core is written in Rust, and available for Python, R and NodeJS.
> 
> The goal of Polars is to provide a lightning fast DataFrame library that:
>
> * Utilizes all available cores on your machine.
> * Optimizes queries to reduce unneeded work/memory allocations.
> * Handles datasets much larger than your available RAM.
> * Provides a consistent and predictable API.
> * Adheres to a strict schema (data-types should be known before running the query).

::: {.small}
```{python}
import polars as pl
pl.__version__
```
:::

## Series

::: {.small}
Just like Pandas, Polars also has a `Series` type used for columns. For a complete list of polars dtypes see [here](https://docs.pola.rs/api/python/stable/reference/datatypes.html).
:::

::: {.columns .xsmall}
::: {.column}
```{python}
pl.Series("ints", [1, 2, 3, 4, 5])
pl.Series("dbls", [1., 2., 3., 4., 5.])
```
:::
::: {.column}
```{python}
pl.Series("bools", [True, False, True, False, True])
pl.Series("strs", ["A", "B", "C", "D", "E"])
```
:::
:::



## Missing values

::: {.small}
In Polars, missing data is represented by the value `null`. This missing value `null` is used for all data types, including numerical types.
:::

::: {.columns .xsmall}
::: {.column width=33%}
```{python}
pl.Series("ints", 
  [1, 2, 3, None])
pl.Series("dbls", 
  [1., 2., 3., None])
```
:::
::: {.column width=33% .fragment}
```{python}
pl.Series("bools", 
  [True, False, True, None])
pl.Series("strs", 
  ["A", "B", "C", None])
```
:::
::: {.column width=33% .fragment}
```{python}
#| error: true
pl.Series("ints", 
  [1, 2, 3, np.nan])
pl.Series("dbls", 
  [1., 2., 3., np.nan])
```
:::
:::


## Missing value checking

Checking for missing values can be done via the `is_null()` method

::: {.columns .xsmall}
::: {.column}
```{python}
pl.Series("ints", 
  [1, 2, 3, None]).is_null()
pl.Series("dbls", 
  [1., 2., 3., None]).is_null()
```
:::
::: {.column}
```{python}
pl.Series("dbls", 
  [1., 2., 3., np.nan]).is_null()
pl.Series("bools", 
  [True, False, True, None]).is_null()
```
:::
:::

## `null` vs `NaN`

::: {.small}
In Polars, `null` and `NaN` are distinct concepts: `null` represents missing data while `NaN` is a valid IEEE 754 floating point value. This distinction matters for aggregation and filtering.
:::

::: {.columns .xsmall}
::: {.column width='50%'}
```{python}
s = pl.Series("x", [1., 2., np.nan, None])
s
```

```{python}
s.is_null()
s.is_nan()
```
:::
::: {.column width='50%' .fragment}
```{python}
s.null_count()
s.sum()
s.mean()
```

```{python}
s.fill_null(0)
s.fill_nan(0)
```
:::
:::


## DataFrames

Data Frames can be constructed in the same way as Pandas, 

```{python}
#| include: false
np.random.seed(42)
```

::: {.xsmall}
```{python}
df = pl.DataFrame(
  {
    "name":   ["anna","bob","carol", "dave", "erin"],
    "id":     np.random.randint(100, 999, 5),
    "weight": np.random.normal(70, 20, 5),
    "height": np.random.normal(170, 15, 5),
    "date":   pd.date_range(start='2/1/2025', periods=5, freq='D')
  },
  schema_overrides = {"id": pl.UInt16, "weight": pl.Float32}
)
df
```
:::


## Expressions

Polars makes use of lazy evaluation to improve its flexibility and computational performance.

::: {.small}
```{python}
bmi_expr = pl.col("weight") / ((pl.col("height")/100) ** 2)
bmi_expr
```
:::

<br/>

This represents a potential computation that can be executed later. Much of the power of Polars comes from the ability to chain together / compose these expressions.

## Contexts

Contexts are the environments in which expressions are evaluated - examples of common contexts include: `select`, `with_columns`, `filter`, and `group_by`.

::: {.columns .xsmall}
::: {.column width=33%}
```{python}
df.select(bmi = bmi_expr)
```
:::
::: {.column width=66%}
```{python}
df.with_columns(bmi = bmi_expr)
```
:::
:::

## `filter()`

::: {.columns .xxsmall}
::: {.column}
```{python}
df.filter(
  pl.col("height") > 160,
  pl.col("id") < 500
)
```
:::
::: {.column}
```{python}
df.filter(
  (pl.col("height") > 160) | 
  (pl.col("id") < 500)
)
```
:::
:::

## `group_by()` & `agg()`

::: {.xsmall}
```{python}
df.group_by(
  id_range = pl.col("id") - pl.col("id") % 100
).agg(
  pl.len(),
  pl.col("name"),
  bmi_expr.alias("bmi"),
  pl.col("weight", "height").mean().name.prefix("avg_"),
  med_height = pl.col("height").median()
)
```
:::


## More expression expansion

::: {.small}
```{python}
num_cols = pl.col(pl.Float64, pl.Float32)

df.with_columns(
  ((num_cols - num_cols.mean())/num_cols.std()).name.suffix("_std")
)
```
:::

## `sort()` & `sort_by()`

`sort()` sorts a DataFrame by one or more columns. `sort_by()` is the expression-level equivalent for use within contexts.

:::: {.columns .xsmall}
::: {.column width='50%'}
```{python}
df.sort("weight")
df.sort("weight", descending=True)
```
:::

::: {.column width='50%' .fragment}
```{python}
df.sort("id", "height")
```

```{python}
df.with_columns(
  rank = pl.col("weight")
    .rank()
    .over("id")
    .sort_by("weight")
)
```
:::
::::


## Pivoting - long to wide

```{python}
#| include: false
df_long = pl.DataFrame({
  "country": ["A","A","A","A","B","B","B","B","C","C","C","C"],
  "year":    [1999,1999,2000,2000,1999,1999,2000,2000,1999,1999,2000,2000],
  "type":    ["cases","pop","cases","pop","cases","pop","cases","pop","cases","pop","cases","pop"],
  "count":   ["0.7K", "19M", "2K", "20M", "37K", "172M", "80K", "174M", "212K", "1T", "213K", "1T"]
})
```

:::: {.columns .small}
::: {.column width='50%'}
```{python}
df_long
```
:::

::: {.column width='50%' .fragment}
```{python}
df_wide = df_long.pivot(
  on="type",
  index=["country","year"],
  values="count"
)
df_wide
```
:::
::::

## Wide to long (unpivot)

::: {.small}
In polars, `melt()` has been renamed to `unpivot()`.
:::

```{python}
#| include: false
df_wide2 = pl.DataFrame({
  "country": ["A","B","C"],
  "1999":    ["0.7K","37K","212K"],
  "2000":    ["2K","80K","213K"]
})
```

:::: {.columns .small}
::: {.column width='50%'}
```{python}
df_wide2
```
:::

::: {.column width='50%' .fragment}
```{python}
df_wide2.unpivot(
  on=["1999", "2000"],
  index="country",
  variable_name="year",
  value_name="cases"
)
```
:::
::::


## pandas vs polars

::: {.small}

**pandas** is a good choice when:

* Working interactively with small to medium datasets
* Using libraries that expect pandas DataFrames (e.g. scikit-learn, statsmodels, seaborn)
* You need the extensive ecosystem of pandas-compatible tools

**polars** is a good choice when:

* Working with large datasets that benefit from multi-core processing
* Building data pipelines where lazy evaluation can optimize query plans
* You want strict type checking and predictable behavior (no index surprises)

In practice, the two are easy to convert between via `df.to_pandas()` and `pl.from_pandas()`.

:::

## NYC Taxi Data

::: {.small}
```{python}
df = pl.scan_parquet(
  "~/Scratch/nyctaxi/fix/*_fix.parquet"
)
df
```
:::

. . .

::: {.small}

```{python}
df.select(pl.len()).collect()
```
:::

::: {.aside}
These data are available on `rstudio.stat.duke.edu` in the `/data/nyctaxi` directory.
:::

## Query plan

::: {.small}
```{python}
print( df.select(pl.len()).explain() )
```
:::

## Large lazy queries

::: {.xsmall}
```{python}
zone_lookup = pl.read_csv(
  "https://d37ci6vzurychx.cloudfront.net/misc/taxi_zone_lookup.csv"
).rename(
  {"LocationID": "pickup_zone"}
)
```
:::

. . .

::: {.xsmall}
```{python}
query = (
  df
  .filter(pl.col("trip_distance") > 0)
  .rename({"PULocationID": "pickup_zone"})
  .group_by("pickup_zone")
  .agg(
    num_rides = pl.len(),
    avg_fare_per_mile = (pl.col("fare_amount") / pl.col("trip_distance")).mean().round(2)
  ).join(
    zone_lookup.lazy(),
    on = "pickup_zone",
    how = "left"
  )
  .sort("pickup_zone")
)
```
:::

## Plan {.scrollable}

::: {.xsmall}
```{python}
query
```
:::

## Result

::: {.xsmall}
```{python}
query.collect()
```
:::


## pandas equivalent

::: {.xsmall}
```{python}
# pyarrow is required for reading parquet files

zone_lookup_pd = pd.read_csv(
  "https://d37ci6vzurychx.cloudfront.net/misc/taxi_zone_lookup.csv"
).rename(columns={"LocationID": "pickup_zone"})

def pd_query():
    df = pd.read_parquet("~/Scratch/nyctaxi/fix/")
    filtered = (
      df[df["trip_distance"] > 0]
      .rename(columns={"PULocationID": "pickup_zone"})
      .assign(fare_per_mile=lambda x: x["fare_amount"] / x["trip_distance"])
    )
    return (
      filtered
      .groupby("pickup_zone")
      .agg(
        num_rides=("fare_per_mile", "size"),
        avg_fare_per_mile=("fare_per_mile", "mean")
      )
      .round({"avg_fare_per_mile": 2})
      .reset_index()
      .merge(zone_lookup_pd, on="pickup_zone", how="left")
      .sort_values("pickup_zone")
    )
```
:::


## Performance

#### polars

::: {.small}
```{python}
#| eval: false
%timeit query.collect()
```
```
1.10 s ± 72 ms per loop (mean ± std. dev. of 3 runs, 1 loop each)
```
:::

<br/>

#### pandas

::: {.small}
```{python}
#| eval: false
%timeit pd_query()
```
```
150.8 s ± 5.9 s per loop (mean ± std. dev. of 3 runs, 1 loop each)
```
:::
